---
title: "Tapirus - Cerrado"
author: "Tati Micheletti"
date: "21 January 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(rdrop2)
 token <- drop_auth()
```

# Project Description

The present project aimed at ...

# Hematology and Biochemestry

## Analysis summary

The following analysis were performed for the results to be comparable to the `"Health Assessment of Wild Lowland Tapirs"` *[Journal of Wildlife Disease, 2014]*.

1. Shapiro-Wilk to test Normality of Samples
  + 1a. Non-normal try transform [log(e)] --> test SW again
    -  1a1. Non-normal again --> NON PARAMETRIC
  + 1b. Normal, pass to 2.
2. Levene's test of Variance homogeneity
  + 2a. Non-homogeneous try transform [log(e)] --> test Levene again
    -  2a1. Non-homogeneous --> NON PARAMETRIC
  + 2b. Normal, use --> PARAMETRIC

FOR NON-PARAMETRIC SAMPLES: `Kruskal-Wallis` test for more than 2 classes or `Mann-Whitney U` test for 2 classes
FOR PARAMETRIC SAMPLES: `ANOVA` for more than 2 classes or `T-TEST` for 2 classes

The groups were hierarchically designed as: Biome > Sex > Age. I.e. Pantanal Males Juv x Pantanal Males Adults and only the variables that were *not signifficant* for each group were compared on the next level.

## Coding

### 1. Load Data

Loading the data to work with from a `.csv` file.
```{r load}
 hembio <- drop_read_csv("dataTapirus/hematobioqFull.csv", dtoken = token) #If it doesn't work, turn on the VPN
# hembio <- read.csv("C:/Users/Tati/Dropbox/dataTapirus/hematobioqFull.csv") # loading from the computer
#hembio <- test[ , order(names(test))] # Don't remember why this... =/
```
**SIDE NOTE: If the dataloading fails, turn on the VPN!.**

Here is the final complete table for biochemestry and hemathology (Mata Atlantica, Pantanal and Cerrado):
```{r Table HEMBIO, eval=TRUE, echo=FALSE}
knitr::kable(as.data.table(hembio))
```

### 2. Shapiro-Wilk test

This test was performed to observe if the samples have a relatively normal distribution. Each answer was stored on a table:
```{r shapiro-test}
fDataCol <- 5 #First data column
sw.test <- list()
for (i in fDataCol:ncol(hembio)){
sw.test[[i]] <- shapiro.test(hembio[,i])
sw.test[[i]] <- sw.test[[i]]$p.value}

Variables <- colnames(hembio[,fDataCol:ncol(hembio)])
sw.pvalue <- as.numeric(unlist(sw.test))
sw.results <- data.frame(Variables, sw.pvalue)
sw.results$Normality <- NA
sw.results <- as.data.table(sw.results)
for (i in 1:nrow(sw.results)){
   if (sw.results$sw.pvalue[i]>0.05){
       sw.results$Normality[i] <- "NORMAL" }}
```

Variables that have a normal distribution are marked as `NORMAL` in the `Normality` column.

```{r primary results shapiro, echo=FALSE, eval=TRUE}
knitr::kable(sw.results)
nat.normal <- as.character(sw.results$Variables[!is.na(sw.results$Normality)])
```

Now, variables that are not normal (described as `NA` in the `Normality` column) have to be log(e) - natural logarithm - transforemed and tested again:

```{r variables transformation}
var.transform <- as.character(sw.results$Variables[is.na(sw.results$Normality)])
hembio.log <- cbind(hembio[,c("ID","Biome","Sex","Age")],log(hembio[,var.transform]))
```

```{r variables transformed, echo=FALSE, eval=TRUE}
knitr::kable(as.data.table(hembio.log))
```

Now we apply again the Shapiro-Wilk test on the transformed variables and see if any of them can be used:
```{r shapiro-test2}
sw.test2 <- list()
for (i in fDataCol:ncol(hembio.log)){
sw.test2[[i]] <- shapiro.test(hembio.log[,i])
sw.test2[[i]] <- sw.test2[[i]]$p.value}

Variables <- colnames(hembio.log[,fDataCol:ncol(hembio.log)])
sw.pvalue <- as.numeric(unlist(sw.test2))
sw.results2 <- data.frame(Variables, sw.pvalue)
sw.results2$Normality <- NA
sw.results2 <- as.data.table(sw.results2)
for (i in 1:nrow(sw.results2)){
   if (is.finite(sw.results2$sw.pvalue[i])&sw.results2$sw.pvalue[i]>0.05){
       sw.results2$Normality[i] <- "NORMAL" }}
```

Variables that presented a normal distribution after transformation are marked as `NORMAL` in the `Normality` column.

```{r secondary results shapiro, echo=FALSE, eval=TRUE}
knitr::kable(sw.results2)
non.normal <- as.character(sw.results2$Variables[is.na(sw.results2$Normality)])
normal.trans <- as.character(sw.results2$Variables[!is.na(sw.results2$Normality)])
```

List of variables that will naturally be treated as `NON-NORMAL`, and therefore will be analyzed with `NON-PARAMETRIC` tests:
```{r non parametric, echo=FALSE, eval=TRUE}
sw.results2$Normality[is.na(sw.results2$Normality)] <- "NON-NORMAL"
knitr::kable(sw.results2[sw.results2$Normality=="NON-NORMAL"])
```
**NOTE: It is important to not that more variables will be added to this list after the homogeneity test.**  

### 3. Levene's test

After the normality test, the Levene's test for homogeneity of variance across groups needs to be performed on the following variables: 
```{r homogen, echo=FALSE, eval=TRUE}
library(car)
normal <- c(nat.normal, normal.trans)
hembio.homog <- cbind(hembio[,1:(fDataCol-1)],hembio[normal])
knitr::kable(as.data.table(hembio.homog))
homo.test <- colnames(hembio.homog[5:ncol(hembio.homog)])
```

This test was performed to observe if the samples present a homogeneity of variances. Each answer was stored on a table:
```{r levene}
source(file.path(getwd(),"R functions/Levene.R"))

  pant.m <- hembio.homog[hembio.homog$Biome=="Pantanal"&hembio.homog$Sex=="Male",]
  pant.f <- hembio.homog[hembio.homog$Biome=="Pantanal"&hembio.homog$Sex=="Female",]
  atl <- hembio.homog[hembio.homog$Biome=="Atlantic",] #Can't be performed at Age level because have only one juvenile, performed at Sex level
  cerr.m <- hembio.homog[hembio.homog$Biome=="Cerrado"&hembio.homog$Sex=="Male",]
  cerr.f <- hembio.homog[hembio.homog$Biome=="Cerrado"&hembio.homog$Sex=="Female",]

lv.pant.m <- levene.p(pant.m,fDataCol, "Age")
lv.pant.f <- levene.p(pant.f,fDataCol, "Age")
lv.atl <- levene.p(atl,fDataCol, "Sex") 
lv.cerr.m <- levene.p(cerr.m,fDataCol, "Age")
lv.cerr.f <- levene.p(cerr.f,fDataCol, "Age")
```

```{r homogen final tables, echo=FALSE, eval=TRUE}
knitr::kable(lv.pant.m)
knitr::kable(lv.pant.f)
knitr::kable(lv.atl)
knitr::kable(lv.cerr.m)
knitr::kable(lv.cerr.f)
```

The following variables were not homogeneous for the following groups:
PANTANAL: MALES: `Plaquetas`, `LeucocitosTotais` and `LactatoDesigrogenase`
PANTANAL: FEMALES: `ColesterolFracionadoHDL` and `LactatoDesigrogenase`
ATLANTIC: BOTH SEXES: `ColesterolFracionadoLDL`,  `ColesterolFracionadoVLDL` and `LactatoDesigrogenase`
CERRADO: MALES: `Hemoglobina` and `Plaquetas`
CERRADO: FEMALES: `Hemoglobina`, `Plaquetas` and `ColesterolFracionadoVLDL`

```{r vector non homogenous}
non.homo <- c("Plaquetas", "LeucocitosTotais", "LactatoDesigrogenase", "ColesterolFracionadoHDL", "ColesterolFracionadoLDL", "ColesterolFracionadoVLDL", "Hemoglobina", "Plaquetas")
```

Therefore, these will be tested using `NON-PARAMETRIC` statistics, since at least in one group these are not homogenous.

Now it is necessary to check which variables have enough samples to perform a comparative mean test. First, we count how many values different than `NA` we have for each variable in each group:

```{r testing, echo=FALSE, eval=TRUE}
library(dplyr)
library(reshape2)
hembio.dt <- as.data.table(hembio[,!colnames(hembio)=="ID"])
binary.dt <- as.data.frame(hembio.dt)

# Converting the table to binary (o/1) so I know which variables 
# don't have enough data to compare
for (i in 1:nrow(binary.dt)){
  for (j in 4:ncol(binary.dt)){
    if (!is.na(binary.dt[i,j]))
        binary.dt[i,j] <- 1
  }
}
binary.dt$BSA <- paste(binary.dt$Biome, binary.dt$Sex, binary.dt$Age, sep = ":")
binary.dt <- binary.dt[,4:ncol(binary.dt)]
count.df <- binary.dt %>% 
  melt(id.var="BSA") %>%
  filter(!is.na(value)) %>%  # Remove NA
  group_by(BSA, variable, value) %>%
  summarise(count=n()) %>% 
  group_by(BSA, variable)
count.df <- as.data.frame(count.df)
count.df <- tidyr::separate(data = count.df, col = BSA, sep = ":", into = c("Biome","Sex","Age"))
count.df <- as.data.table(count.df)
```

```{r count.df, echo=TRUE, eval=TRUE}
knitr::kable(count.df)
```

We need to have >2 for biome and sex, >3 for ages, otherwise we can't compare the means
```{r}
meanTestBiome <- apply(count.df, 1, FUN = function(x){
  meanTestBiome <- ifelse(as.numeric(x["count"])>2,"YES",NA)
  return(meanTestBiome)
})

meanTestSex <- apply(count.df, 1, FUN = function(x){
  meanTestSex <- ifelse(as.numeric(x["count"])>2,"YES",NA)
  return(meanTestSex)
})

meanTestAges <- apply(count.df, 1, FUN = function(x){
  meanTestAges <- ifelse(as.numeric(x["count"])>3,"YES",NA)
  return(meanTestAges)
})

count.df <- cbind(count.df,meanTestBiome,meanTestSex,meanTestAges)
```

For the meanTest that have `YES`, we can run either a T-Test or an ANOVA (or their comparable non-parametric).
To identify which one will be done fore each variable/sex/age/biome, we will create a complete table with all variables, sex, biomes and ages
```{r expanded.grid}
var.excl <- c("Sex", "Biome", "Age")
variables <- as.character(colnames(hembio.dt))
variables <- c(variables[!variables %in% var.excl])
ages <- as.character(unique(hembio.dt$Age))
biome <- as.character(unique(hembio.dt$Biome))
sex <- as.character(unique(hembio.dt$Sex))
completeTable <- expand.grid(variable = variables, Age = ages, Sex = sex, Biome = biome) %>%
  data.table()
```

Now we match the complete table and the one with the counts:
```{r join.tables}
joinTable <- merge(count.df,completeTable, all = TRUE) %>%
    .[,-5]

```

First: for each variable, I have to check for each age of each sex of each Biome, if we have at least 1 count.
We subset this table per Biome and Sex, to observe the variables for each age and for sex:
```{r subset, echo=FALSE, eval=TRUE}
At.F.a <- joinTable[Biome=="Atlantic"&Sex=="Female"&Age=="Adult"]
At.F.s <- joinTable[Biome=="Atlantic"&Sex=="Female"&Age=="Sub-adult"]
At.F.j <- joinTable[Biome=="Atlantic"&Sex=="Female"&Age=="Juvenile"]

At.M.a <- joinTable[Biome=="Atlantic"&Sex=="Male"&Age=="Adult"]
At.M.s <- joinTable[Biome=="Atlantic"&Sex=="Male"&Age=="Sub-adult"]
At.M.j <- joinTable[Biome=="Atlantic"&Sex=="Male"&Age=="Juvenile"]

Pt.F.a <- joinTable[Biome=="Pantanal"&Sex=="Female"&Age=="Adult"]
Pt.F.s <- joinTable[Biome=="Pantanal"&Sex=="Female"&Age=="Sub-adult"]
Pt.F.j <- joinTable[Biome=="Pantanal"&Sex=="Female"&Age=="Juvenile"]

Pt.M.a <- joinTable[Biome=="Pantanal"&Sex=="Male"&Age=="Adult"]
Pt.M.s <- joinTable[Biome=="Pantanal"&Sex=="Male"&Age=="Sub-adult"]
Pt.M.j <- joinTable[Biome=="Pantanal"&Sex=="Male"&Age=="Juvenile"]

Cr.F.a <- joinTable[Biome=="Cerrado"&Sex=="Female"&Age=="Adult"]
Cr.F.s <- joinTable[Biome=="Cerrado"&Sex=="Female"&Age=="Sub-adult"]
Cr.F.j <- joinTable[Biome=="Cerrado"&Sex=="Female"&Age=="Juvenile"]

Cr.M.a <- joinTable[Biome=="Cerrado"&Sex=="Male"&Age=="Adult"]
Cr.M.s <- joinTable[Biome=="Cerrado"&Sex=="Male"&Age=="Sub-adult"]
Cr.M.j <- joinTable[Biome=="Cerrado"&Sex=="Male"&Age=="Juvenile"]

tableList <- list("Atlantic" = list("Female" = list("Adult" = At.F.a,
                                                "Sub-adult" = At.F.s,
                                                "Juvenile" = At.F.j), 
                                  "Male" = list("Adult" = At.M.a,
                                                "Sub-adult" = At.M.s,
                                                "Juvenile" = At.M.j)), 
                  "Pantanal" = list("Female" = list("Adult" = Pt.F.a,
                                                "Sub-adult" = Pt.F.s,
                                                "Juvenile" = Pt.F.j), 
                                  "Male" = list("Adult" = Pt.M.a,
                                                "Sub-adult" = Pt.M.s,
                                                "Juvenile" = Pt.M.j)), 
                  "Cerrado" = list("Female" = list("Adult" = Cr.F.a,
                                                "Sub-adult" = Cr.F.s,
                                                "Juvenile" = Cr.F.j), 
                                  "Male" = list("Adult" = Cr.M.a,
                                                "Sub-adult" = Cr.M.s,
                                                "Juvenile" = Cr.M.j)))
```

Check for each Age if we can compare for each variable:
``` {r lapplyoflapply}

compSEX <- lapply(biome, function(biome){
  
  tests1 <- lapply(sex, function(sex){

        listAges <- lapply(tableList[[biome]][[sex]], function(x){
          x$Age <- NULL
            return(x)
        })
        
      listSex <- lapply(listAges, function(x){
          nx <- x[,c("Biome","Sex","variable","count")]
              return(nx)
            })

           
        concListSex <- lapply(listSex, function(i) setkey(i, variable))
        merged <- Reduce(function(...) merge(..., all = FALSE), listSex)
        merged$countF <- apply(merged, 1, function(x){
          sum(as.numeric(x["count.x"]), as.numeric(x["count.y"]), as.numeric(x["count"]), na.rm = TRUE)})
        merged <- merged[,-c(2:7, 10)]

return(merged)
        
  })
  
  names(tests1) <- sex
    tests2 <- tests1
        lapply(tests2, function(i) setkey(i, variable))
        biomeTable <- Reduce(function(...) merge(..., all = FALSE), tests2)
        biomeTable$test <- apply(biomeTable, 1, function(x){
           ifelse(as.numeric(x["countF.x"])>2&as.numeric(x["countF.y"])>2,"T",NA)
         })
        biomeTable <- biomeTable[, c(1,8)]
    return(biomeTable)
})

names(compSEX) <- biome

#########################################################################

lapply(compSEX, function(i) setkey(i, variable))
compBIOME <-  Reduce(function(...) merge(..., all = FALSE), compSEX)
colnames(compBIOME) <- c("variable",as.character(biome))
compBIOME$whichGroups <- apply(compBIOME, 1, function(x){
            whichG <- attributes(which(!is.na(x[2:length(x)])))$names %>%
             paste(collapse = ":")
            })
          
          compBIOME$manyGroups <- apply(compBIOME, 1, function(x){
            subX <- x[2:4]
            x <-  sum(!is.na(subX))
            return(x)})
          compBIOME$test <- ifelse(compBIOME$manyGroups<2,NA,ifelse(compBIOME$manyGroups==2,"T","ANOVA"))
          compBIOME <- compBIOME[,-c(2:4,6)]

########################################################################

compAGE <- lapply(biome, function(biome){

  tests1 <- lapply(sex, function(sex){
    
        listAges <- lapply(tableList[[biome]][[sex]],"[[","meanTestAges")
          df <- do.call(cbind.data.frame,listAges)
          df2 <- df
          df2$whichGroups <- apply(df, 1, function(x){
            
            whichG <- attributes(which(!is.na(x)))$names %>%
             paste(collapse = ":")
            })
          
          df2$manyGroups <- apply(df, 1, function(x) sum(!is.na(x)))
          df2$test <- ifelse(df2$manyGroups<2,NA,ifelse(df2$manyGroups==2,"T","ANOVA"))
          df2 <- cbind(variables, df2) %>%
            .[,-c(2:4,6)]
          df2$test <- ifelse(is.na(df2$test),NA,df2$test)
          
          return(df2)
  })
    names(tests1) <- sex
    return(tests1)
})
names(compAGE) <- biome

```

Here are the tests by groups: 

AGE:
```{r table compAGE, echo=FALSE, eval=TRUE}
knitr::kable(as.data.table(compAGE))
```

SEX:
```{r table compSEX, echo=FALSE, eval=TRUE}
knitr::kable(as.data.table(compSEX))
```

BIOME:
```{r table compBIOME, echo=FALSE, eval=TRUE}
knitr::kable(as.data.table(compBIOME))
```

The following table presents a summary of the analysis to be performed for each variable:
```{r final.table, echo=FALSE, eval=TRUE}
test_groups <- data.table(PARAM = rep(c("PARAMETRIC", "NON-PARAMETRIC"), each = 2), GROUP = c(2,3,2,3), TEST = c("T-Test", "ANOVA", "Mann-WhitneyU","Kruskal-Wallis"))
all.var <- colnames(hembio[,c(fDataCol:ncol(hembio))])
final.table <- data.frame(VARIABLES = all.var, NORMALITY = NA, TRANSFORMATION = NA, HOMOGENEITY = NA, TEST = NA)
for (i in 1:nrow(final.table)){
  if (final.table[i,"VARIABLES"] %in% nat.normal)
          final.table[i,"NORMALITY"] <- "NORMAL" else 
          final.table[i,"NORMALITY"] <- "NON-NORMAL"
  if (final.table[i,"VARIABLES"] %in% normal.trans) 
          final.table[i,"TRANSFORMATION"] <- "NORMAL" else
      if (final.table[i,"NORMALITY"]=="NORMAL") 
              final.table[i,"TRANSFORMATION"] <- "NA" else
              final.table[i,"TRANSFORMATION"] <- "NON-NORMAL"
  if (!(final.table[i,"VARIABLES"] %in% homo.test))
          final.table[i,"HOMOGENEITY"]  <- "NA" else
      if (final.table[i,"VARIABLES"] %in% non.homo) 
              final.table[i,"HOMOGENEITY"] <- "NON-HOMOGENEOUS" else 
              final.table[i,"HOMOGENEITY"] <- "HOMOGENEOUS"
  if (!(final.table[i,"TRANSFORMATION"] == "NON-NORMAL") & final.table[i,"HOMOGENEITY"] == "HOMOGENEOUS")
              final.table[i,"TEST"] <- "PARAMETRIC" else
              final.table[i,"TEST"] <- "NON-PARAMETRIC"
}

final.table <- data.table(final.table)
```

```{r table final.table, echo=FALSE, eval=TRUE}
knitr::kable(as.data.table(final.table))
```

Now, for each variable I just need to merge the last table (`final.table`) with the AGE, SEX and BIOME, to be able to automatically loop through the variables and perform the comparisons. I will start from AGE, then SEX, and at last BIOME. It is important to note that significant differences in any levels will not allow for comparison on the superior levels. 

```{r analysis.table, echo=FALSE, eval=TRUE}
#----------------- AGE ---------------

AGE.table <- lapply(biome, function(biome){
  TEST <- lapply(sex, function(sex){
      TEST <- apply(compAGE[[biome]][[sex]], 1, function(rows){
        
                  line <- which(final.table[,VARIABLES]==rows["variables"])
                  tst <- final.table[line,TEST]
                  
              TEST <- ifelse(rows["test"]=="T",
                        ifelse(tst=="PARAMETRIC","T-TEST","Mann-WhitneyU"),
                          ifelse(rows["test"]=="ANOVA",
                            ifelse(tst=="PARAMETRIC","ANOVA","Kruskal-Wallis"),
                              NA))
          return(TEST)
      })
        return(TEST)
    })
  names(TEST) <- sex
      return(TEST)
  })
names(AGE.table) <- biome

AGE <- lapply(biome, function(biome){
    newAGE <- lapply(sex, function(sex){
      statTest <- cbind(compAGE[[biome]][[sex]], statTest = AGE.table[[biome]][[sex]])
      return(statTest)
  })
    names(newAGE) <- sex
    return(newAGE)
})
names(AGE) <- biome

#----------------- SEX ---------------

SEX.table <- lapply(biome, function(biome){
      TEST <- apply(compSEX[[biome]], 1, function(rows){

               line <- which(final.table[,VARIABLES]==rows["variable"])
               tst <- final.table[line,TEST]
                  
              TEST <- ifelse(rows["test"]=="T",
                        ifelse(tst=="PARAMETRIC","T-TEST","Mann-WhitneyU"),
                              NA)
        return(TEST)
    })
      return(TEST)
  })
names(SEX.table) <- biome

SEX <- lapply(biome, function(biome){
      statTest <- cbind(compSEX[[biome]], statTest = SEX.table[[biome]])
      return(statTest)
  })

names(SEX) <- biome

#----------------- BIOME ---------------

BIOME.table <- apply(compBIOME, 1, function(rows){

            line <- which(final.table[,VARIABLES]==rows["variable"])
            tst <- final.table[line,TEST]
                  
              TEST <- ifelse(rows["test"]=="T",
                        ifelse(tst=="PARAMETRIC","T-TEST","Mann-WhitneyU"),
                          ifelse(rows["test"]=="ANOVA",
                            ifelse(tst=="PARAMETRIC","ANOVA","Kruskal-Wallis"),
                              NA))
          return(TEST)
      })

BIOME <- cbind(compBIOME, statTest = BIOME.table)

```

```{r tables for stats, echo=FALSE, eval=TRUE}
knitr::kable(as.data.table(AGE))
knitr::kable(as.data.table(SEX))
knitr::kable(as.data.table(BIOME))
```

As we have the final tables indicating which tests need to be performed, we built a function to perform these tests. The function is on a separate `R script` on the folder **R functions** and is automatically sourced with the code below.  

*NOTE:* **It is necessary to run the code below in the given order as the first function creates a table to exclude variables that are significant for age from the sex analysis and so forth.**

```{r stats, echo=FALSE, eval=TRUE}

source(file.path(getwd(),"R functions/statTests.R"))

statAGE <- statTests(listTests = AGE, 
                     testsPerVariable = final.table, 
                     originalDT = hembio, 
                     logDT = hembio.log, 
                     biome = biome, 
                     sex = sex)

statSEX <- statTests(listTests = SEX, 
                     testsPerVariable = final.table, 
                     originalDT = hembio, 
                     logDT = hembio.log, 
                     biome = biome, 
                     sex = sex,
                     exclude = outSEX.list)

# Still need to finish the statTest.R script for the statBiome... Already started, but need to finish

statBIOME <- statTests(listTests = BIOME, 
                     testsPerVariable = final.table, 
                     originalDT = hembio, 
                     logDT = hembio.log, 
                     biome = biome, 
                     sex = sex,
                     exclude = outBIOME.list)

```

```{r tables with p.value, echo=FALSE, eval=TRUE}
knitr::kable(as.data.table(statAGE))
knitr::kable(as.data.table(statSEX))
knitr::kable(as.data.table(statBIOME))
```

After finishing above, need to make posthoc of ANOVAS and K-W, but then it is finished. 
