---
title: "Tapirus terrestris: Health analysis - MICROBIOLOGY"
author: "Tati Micheletti"
date: "May 15, 2018"
output: html_document
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "Results") })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(data.table)
library(reproducible)
library(rcompanion)
library(betapart)
```

# Project Description

The present project aimed at understanding the general health of Tapir (**Tapirus terrestris**) populations on three different biomes.

# Microbiology

## Analysis summary
bacteria: only prevalence (X2) and Jaccard

1. Prevalences --> % positives on the population. Use the positives in the total tested. (For each bacteria, positives/total ind. tested) Ex. 
A - bac1, bac2
B - Negative
C - bac1
D - bac1
E - Negative

Prevalence bac1: 3/5 = 0.6 or 60%
Prevalence bac2: 1/5 = 0.2 or 20%

2. Jaccard similarity coefficient for bacterias for cavities. Sj=1, same bio profile, Sj=0, different
     [Sj = j/(a+b-j)], j = number of species of bacteria that occurred in both the tapir populations for each cavity in the AF, PA (and CE?) study sites; a = number of species of bacteria that occurred in the PA population; b = number of species of bacteria that occurred in the AF population. AF and PA --> Sj = 0.125
3. x2 (chi-square) test for prevalence in different bacterias in study sites --> Not significant for the bacts found in both PT and MA

## Coding

### 1. Load Data
Loading the data to work with from a `.csv` file.
```{r load}
tryCatch({
  prevalence <- data.table::fread(file.path(getwd(), "dataTapirus/Microbiology/microbiology.csv")
                                  )}, error = function(e){
  stop("File 'microbiology.csv' not found. You likely do not have access to the data. Please contact the webpage author for more information.")
}
)

```

### 2. Prevalence of bacterial species in the three populations (independently of the cavity) 
```{r prevalence}

prevalTable <- prevalence[, .(freq = sum(frequency)), by = c("biome", "species")]
prevalTableTotal <- prevalTable[, .(totalCounts = sum(freq)), by = "biome"]
data.table::rbindlist(lapply(X = unique(prevalTable$biome), FUN = function(bio){
  neg <- prevalTableTotal[biome == bio, totalCounts]
  prevalTable[biome == bio, negative := neg - freq]
  })
)

cols <- unique(prevalTable$species) %>%
  .[. != "NEGATIVE"]

biomes <- unique(prevalTable$biome)

CI <- data.table::rbindlist(lapply(X = cols, FUN = function(bac){
  biome.ci <- data.table::rbindlist(lapply(X = biomes, FUN = function(biom){
      sb <- prevalTable[biome == biom & species == bac,]
  total <- sb$freq + sb$negative
  positive <- sb$freq
  if (nrow(sb) != 0) {
      CI.calc <- stats::binom.test(x = positive, n = total, p = 0.5,
                                   alternative="two.sided",
                                   conf.level=0.95)
  CI.low <- min(CI.calc$conf.int)
  CI.up <- max(CI.calc$conf.int)
  Ci <- data.table::data.table(bacteria = bac,
                               biome = biom,
                               prevalence = round(CI.calc$estimate, 3)*100,
                               CI = paste0(round(CI.low, 3)*100, " - ", round(CI.up, 3)*100))
  } else {
      Ci <- data.table::data.table(bacteria = bac,
                               biome = biom,
                               prevalence = 0,
                               CI = paste0("0 - 0"))
                               
  }
  return(Ci)

  })
  )
  return(biome.ci)
})
)

CI

write.csv(x = CI, file = file.path(getwd(), "Results/microbioPrevalence.csv"))
```

### 3. X2 of prevalence per bacterial species (sum in cavities) 
```{r x2_total}

x2TotalTable <- prevalence[, c("biome", "species", "frequency")]
x2TotalCast <- data.table::dcast.data.table(data = x2TotalTable, formula = biome ~ species, 
                                            value.var = "frequency", fun.aggregate = sum) %>%
  .[,!"NEGATIVE"]

x2Total <- as.matrix(x2TotalCast[,2:ncol(x2TotalCast)])
rownames(x2Total) <- x2TotalCast$biome
x2TotalComp <- suppressWarnings(rcompanion::pairwiseNominalIndependence(x2Total, compare = "row", fisher = TRUE,
  gtest = FALSE, chisq = FALSE, method = "bonferroni", correct = "none",
  digits = 3, workspace = 1e7, simulate.p.value = TRUE))

x2TotalComp

write.csv(x = x2TotalComp, file = file.path(getwd(), "Results/X2TotalMicrobio.csv"))
```

Here we see that considering the bacterial flora independently of the cavity, all locations are different. However, we do have a lot of data (lots of bacteria species), which means we might be inflating the results (seeing difference when there is not much) Therefore, we should also make a X2 for each location and EACH bacteria. But first, I will calculate the prevalence in each cavity. Then we go for the X2 test.

```{r prev_eachBac}
prevalTableCav <- prevalence[biome == "CERRADO", .(freq = sum(frequency)), by = c("biome", "species", "cavities")]
prevalTableTotalCerr <- prevalTableCav[, .(totalCounts = sum(freq)), by = "cavities"]

data.table::rbindlist(lapply(X = unique(prevalTableCav$cavities), FUN = function(cav){
  neg <- prevalTableTotalCerr[cavities == cav, totalCounts]
  prevalTableCav[cavities == cav, negative := neg - freq]
  })
)

cols <- unique(prevalTableCav$species) %>%
  .[. != "NEGATIVE"]

cavs <- unique(prevalTableCav$cavities)

CI.cav <- data.table::rbindlist(lapply(X = cols, FUN = function(bac){
  cavity.ci <- data.table::rbindlist(lapply(X = cavs, FUN = function(cav){
      sb <- prevalTableCav[cavities == cav & species == bac,]
  total <- sb$freq + sb$negative
  positive <- sb$freq
  if (nrow(sb) != 0) {
      CI.calc <- stats::binom.test(x = positive, n = total, p = 0.5,
                                   alternative="two.sided",
                                   conf.level=0.95)
  CI.low <- min(CI.calc$conf.int)
  CI.up <- max(CI.calc$conf.int)
  Ci <- data.table::data.table(bacteria = bac,
                               cavity = cav,
                               prevalence = round(CI.calc$estimate, 3)*100,
                               CI = paste0(round(CI.low, 3)*100, " - ", round(CI.up, 3)*100))
  } else {
      Ci <- data.table::data.table(bacteria = bac,
                                cavity = cav,
                               prevalence = 0,
                               CI = paste0("0 - 0"))
                               
  }
  return(Ci)

  })
  )
  return(cavity.ci)
})
)

CI.cav

write.csv(x = CI.cav, file = file.path(getwd(), "Results/microbioPrevalenceCavities.csv"))

```


```{r each_bac}
x2each <- data.table::rbindlist(lapply(X = unique(prevalTable[species != "NEGATIVE", species]), FUN = function(bac){
  sb <- prevalTable[species == bac, ]
  if (!"ATLANTIC" %in% unique(sb$biome)){
    rowAT <- data.table::data.table(biome = "ATLANTIC",
                                    species = bac,
                                    freq = 0,
                                    negative = prevalTableTotal[biome == "ATLANTIC", totalCounts])
    sb <- rbind(sb, rowAT)
  }
    if (!"CERRADO" %in% unique(sb$biome)){
    rowCE <- data.table::data.table(biome = "CERRADO",
                                    species = bac,
                                    freq = 0,
                                    negative = prevalTableTotal[biome == "CERRADO", totalCounts])
    sb <- rbind(sb, rowCE)
    }
    if (!"PANTANAL" %in% unique(sb$biome)){
    rowPT <- data.table::data.table(biome = "PANTANAL",
                                    species = bac,
                                    freq = 0,
                                    negative = prevalTableTotal[biome == "PANTANAL", totalCounts])
    sb <- rbind(sb, rowPT)
  }
  matrix <- as.matrix(sb[, c("freq", "negative")])
    rownames(matrix) <- sb$biome
  x2 <- rcompanion::pairwiseNominalIndependence(matrix, compare = "row", fisher = TRUE,
  gtest = FALSE, chisq = TRUE, method = "bonferroni", correct = "none",
  digits = 3)
  x2$species <- bac
  return(suppressWarnings(x2))
})
)

x2each

write.csv(x = x2each, file = file.path(getwd(), "Results/X2EachMicrobio.csv"))
```


### 5. Prevalence calculation per bacterial species per cavity
```{r per_cavity}
listX2 <- lapply(X = unique(prevalence$species[prevalence$species != "NEGATIVE"]), FUN = function(sp){
  assign(x = sp, value = prevalence[species == sp, c("biome", "cavities", "frequency")])
  assign(x = sp, value = data.table::dcast(get(sp), biome ~ cavities, value.var = "frequency"))
  if (nrow(get(sp)) == 3){
    eval(parse(text = paste0("attributes(", sp, ")[['species']] <- sp")))
    return(get(sp))
  }
})

listX2 <- listX2[vapply(listX2, Negate(is.null), NA)]
listNames <- unlist(lapply(X = listX2, FUN = function(species){
 listName <- attributes(species)$species
  return(listName)
})
)
names(listX2) <- listNames

toExclude <- c("urethralSwab", "lesionSwab", "eyeSwab") # These were not collected (or compared) in the first paper, not much information
listX2 <- lapply(X = listX2, function(x){
  x[, !toExclude, with = FALSE]
  })

x2 <- lapply(X = listX2, FUN = function(x){
  for (i in names(x)) x[is.na(get(i)), (i) := 0]
  matrix <- as.matrix(x[,2:ncol(x)])
  rownames(matrix) <- x$biome
  x2 <- pairwiseNominalIndependence(matrix, compare = "row", fisher = TRUE,
  gtest = FALSE, chisq = TRUE, method = "bonferroni", correct = "none",
  digits = 3)
  return(suppressWarnings(x2))
})

X2pVal <- rbindlist(lapply(X = names(x2), FUN = function(species){
  x2[[species]]$species <- species
  return(x2[[species]])
})
)
X2pVal

write.csv(x = X2pVal, file = file.path(getwd(), "Results/X2PerCavityMicrobio.csv"))
```

In this table we can see that the p of both Fisher and X2 tests for E. coli are significant (Cerrado being different from Pantanal and MT Atlantica), but as we have to correct because of the low number of samples, they become non-significant. For `staphylococcusAureus` is not significant. 

### 6. Similarity index
Prepare the table for Jaccard: rows are sites, columns are species
```{r Jaccard2}
colsToKeep <- c("biome", "species")
dissimDF <- data.table::dcast(data = prevalence[, colsToKeep, with = FALSE], formula = biome ~ species) %>%
  as.data.frame()
biomes <- dissimDF[, 1]
dissimDF <- dissimDF[, 3:ncol(dissimDF)]
rownames(dissimDF) <- biomes
dissimDF[dissimDF > 1] <- 1
AT.PT <- dissimDF[c("ATLANTIC", "PANTANAL"),]
CE.PT <- dissimDF[c("CERRADO", "PANTANAL"),]
CE.AT <- dissimDF[c("ATLANTIC", "CERRADO"),]

AT.PT.Sorensen <- betapart::beta.multi(x = AT.PT, index.family = "sorensen")
AT.PT.Jaccard <- betapart::beta.multi(x = AT.PT, index.family = "jaccard")

CE.PT.Sorensen <- betapart::beta.multi(x = CE.PT, index.family = "sorensen")
CE.PT.Jaccard <- betapart::beta.multi(x = CE.PT, index.family = "jaccard")

CE.AT.Sorensen <- betapart::beta.multi(x = CE.AT, index.family = "sorensen")
CE.AT.Jaccard <- betapart::beta.multi(x = CE.AT, index.family = "jaccard")

finalSimilarityTable <- data.table::data.table(
  Comparison = c("ATLANTIC:PANTANAL", "CERRADO:PANTANAL", "CERRADO:ATLANTIC"),
  Sorensen = c(AT.PT.Sorensen$beta.SOR, CE.PT.Sorensen$beta.SOR, CE.AT.Sorensen$beta.SOR),
  Jaccard = c(AT.PT.Jaccard$beta.JAC, CE.PT.Jaccard$beta.JAC, CE.AT.Jaccard$beta.JAC))

tdissimDF <- data.table::transpose(dissimDF)
colnames(tdissimDF) <- rownames(dissimDF)
rownames(tdissimDF) <- colnames(dissimDF)


finalSimilarityTable

write.csv(x = finalSimilarityTable, file = file.path(getwd(), "Results/SimilarityMicrobio.csv"))
```

