---
title: "Tapirus - Cerrado"
author: "Tati Micheletti"
date: "21 January 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(rdrop2)
 drop_auth()
 token <- drop_auth()
 saveRDS(token, file = "token.rds")
```

# Project Description

The present project aimed at ...

# Hematology and Biochemestry

## Analysis summary

The following analysis were performed for the results to be comparable to the `"Health Assessment of Wild Lowland Tapirs"` *[Journal of Wildlife Disease, 2014]*.

1. Shapiro-Wilk to test Normality of Samples
  + 1a. Non-normal try transform [log(e)] --> test SW again
    -  1a1. Non-normal again --> NON PARAMETRIC
  + 1b. Normal, pass to 2.
2. Levene's test of Variance homogeneity
  + 2a. Non-homogeneous try transform [log(e)] --> test Levene again
    -  2a1. Non-homogeneous --> NON PARAMETRIC
  + 2b. Normal, use --> PARAMETRIC

FOR NON-PARAMETRIC SAMPLES: `Kruskal-Wallis` test for more than 2 classes or `Mann-Whitney U` test for 2 classes
FOR PARAMETRIC SAMPLES: `ANOVA` for more than 2 classes or `T-TEST` for 2 classes

The groups were hierarchically designed as: Biome > Sex > Age. I.e. Pantanal Males Juv x Pantanal Males Adults and only the variables that were *not signifficant* for each group were compared on the next level.

## Coding

### 1. Load Data

Loading the data to work with from a `.csv` file.
```{r load}
hembio <- drop_read_csv("dataTapirus/hematobioqFull.csv")
hembio <- test[ , order(names(test))]
```
**SIDENOTE: If the dataloading fails, turn on the VPN!.**

Here is the final complete table for biochemestry and hemathology (Mata Atlantica, Pantanal and Cerrado):
```{r Table HEMBIO, eval=TRUE, echo=FALSE}
knitr::kable(as.data.table(hembio))
```

### 2. Shapiro-Wilk test

This test was performed to observe if the samples have a relatively normal distribution. Each answer was stored on a table:
```{r shapiro-test}
sw.test <- list()
for (i in 5:ncol(hembio)){
sw.test[[i]] <- shapiro.test(hembio[,i])
sw.test[[i]] <- sw.test[[i]]$p.value}

Variables <- colnames(hembio[,5:ncol(hembio)])
sw.pvalue <- as.numeric(unlist(sw.test))
sw.results <- data.frame(Variables, sw.pvalue)
sw.results$Normality <- NA
sw.results <- as.data.table(sw.results)
for (i in 1:nrow(sw.results)){
   if (sw.results$sw.pvalue[i]>0.05){
       sw.results$Normality[i] <- "NORMAL" }}
```

Variables that have a normal distribution are marked as `NORMAL` in the `Normality` column.

```{r primary results shapiro, echo=FALSE, eval=TRUE}
knitr::kable(sw.results)
nat.normal <- as.character(sw.results$Variables[!is.na(sw.results$Normality)])
```

Now, variables that are not normal (described as `NA` in the `Normality` column) have to be log(e) - natural logarithm - transforemed and tested again:

```{r variables transformation}
var.transform <- as.character(sw.results$Variables[is.na(sw.results$Normality)])
hembio.log <- cbind(hembio[,1:4],log(hembio[,var.transform]))
```

```{r variables transformed, echo=FALSE, eval=TRUE}
knitr::kable(as.data.table(hembio.log))
```

Now we apply again the Shapiro-Wilk test on the transformed variables and see if any of them can be used:
```{r shapiro-test2}
sw.test2 <- list()
for (i in 5:ncol(hembio.log)){
sw.test2[[i]] <- shapiro.test(hembio.log[,i])
sw.test2[[i]] <- sw.test2[[i]]$p.value}

Variables <- colnames(hembio.log[,5:ncol(hembio.log)])
sw.pvalue <- as.numeric(unlist(sw.test2))
sw.results2 <- data.frame(Variables, sw.pvalue)
sw.results2$Normality <- NA
sw.results2 <- as.data.table(sw.results2)
for (i in 1:nrow(sw.results2)){
   if (is.finite(sw.results2$sw.pvalue[i])&sw.results2$sw.pvalue[i]>0.05){
       sw.results2$Normality[i] <- "NORMAL" }}
```

Variables that presented a normal distribution after transformation are marked as `NORMAL` in the `Normality` column.

```{r secondary results shapiro, echo=FALSE, eval=TRUE}
knitr::kable(sw.results2)
non.normal <- as.character(sw.results2$Variables[is.na(sw.results2$Normality)])
normal.trans <- as.character(sw.results2$Variables[!is.na(sw.results2$Normality)])
```

List of variables that will naturally be treated as `NON-NORMAL`, and therefore will be analyzed with `NON-PARAMETRIC` tests:
```{r non parametric, echo=FALSE, eval=TRUE}
knitr::kable(sw.results2[is.na(sw.results2$Normality)])
```
**NOTE: It is important to not that more variables will be added to this list after the homogeneity test.**  

### 3. Levene's test

After the normality test, the Levene's test for homogeneity of variance across groups needs to be performed on the following variables: 
```{r homogen, echo=FALSE, eval=TRUE}
library(car)
normal <- c(nat.normal, normal.trans)
hembio.homog <- cbind(hembio[,1:4],hembio[normal])
knitr::kable(as.data.table(hembio.homog))
```

This test was performed to observe if the samples present a homogeneity of variances. Each answer was stored on a table:
```{r levene}
source(file.path(getwd(),"R functions/Levene.R"))
source(file.path(getwd(),"R functions/Levene2.R"))

#  pant <- hembio.homog[hembio.homog$Biome=="Pantanal",] #Age was used for grouping: most restrictive comparison
  pant.m <- hembio.homog[hembio.homog$Biome=="Pantanal"&hembio.homog$Sex=="Male",]
  pant.f <- hembio.homog[hembio.homog$Biome=="Pantanal"&hembio.homog$Sex=="Female",]
  atl <- hembio.homog[hembio.homog$Biome=="Atlantic",] #Can't be performed at Age level because have only one juvenile, performed at Sex level
  cerr.m <- hembio.homog[hembio.homog$Biome=="Cerrado"&hembio.homog$Sex=="Male",]
  cerr.f <- hembio.homog[hembio.homog$Biome=="Cerrado"&hembio.homog$Sex=="Female",]

  # Still not working...
  # 
lv.pant.m <- levene.p(pant.m,5)
lv.pant <- t.levene.p(pant,5)
lv.pant.f <- levene.p(pant.f,5)

lv.atl <- levene2.p(atl,5) 

lv.cerr.m <- levene.p(cerr.m,7)
lv.cerr.f <- levene.p(cerr.f,7)
```

```{r homogen fina tables, echo=FALSE, eval=TRUE}
knitr::kable(lv.pant.m)
knitr::kable(lv.pant.f)
knitr::kable(lv.atl)
knitr::kable(lv.cerr.m)
knitr::kable(lv.cerr.f)
```

Only `LeucocitosTotais` and `ColesterolFracionadoHDL` are not homogenous for Males from Pantanal, and females from Pantanal, respectively. These should also be performed as Non-Parametric for all analysis since at least in one group these are not homogenous.

` MAKE A FINAL TABLE HERE WITH NORMALITY AND HOMOGENEITY FOR EACH VARIABLE AND THE RESPECTIVE TESTS TO BE APPLIED AND AT WHICH VARIABLES (BIOME, SEX, AGE or just BIOME, SEX) or a column that shows KW or TEST T FOR EACH ONE AT EACH LEVEL? Think this through...`

Write NORMAL or NON-NORMAL. Do the same for Levene "HOMOGENEOUS" - "NON-HOMOG". The variables that are NORMAL & HOMOGENEOUS, use parametric, the others non parametric, group by variables in col 2:4.





