---
title: "Tapirus - Cerrado"
author: "Tati Micheletti"
date: "21 January 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Project Description

The present project aimed at ...

# Hematology and Biochemestry

## Analysis summary

The following analysis were performed for the results to be comparable to the `"Health Assessment of Wild Lowland Tapirs"` *[Journal of Wildlife Disease, 2014]*.

1. Shapiro-Wilk to test Normality of Samples
 1a. Non-normal try transform [log(e)] --> test SW again
  1a1. Non-normal again --> NON PARAMETRIC
 1b. Normal, pass to 2.
2. Levene's test of Variance homogeneity
 2a. Non-homogeneous try transform [log(e)] --> test Levene again
  2a1. Non-homogeneous --> NON PARAMETRIC
 2b. Normal, use --> PARAMETRIC

The groups were hierarchically designed as: Biome > Sex > Age. I.e. Pantanal Males Juv x Pantanal Males Adults and only the variables that were *not signifficant* for each group were compared on the next level.

## Coding

### 1. Load Data

Loading the data to work with from a `.csv` file.
```{r load}
library(data.table)
hembio <- fread("hematobioqFull.csv")
hembio <- as.data.frame(hembio) # Easier to perform the Shapiro-test
```

### 2. Shapiro-Wilk test

This test was performed to observe if the samples have a relatively normal distribution. Each answer was stored on a table:
```{r shapiro-test}
sw.test <- list()
for (i in 5:ncol(hembio)){
sw.test[[i]] <- shapiro.test(hembio[,i])
sw.test[[i]] <- sw.test[[i]]$p.value
}
Variables <- colnames(hembio[,5:ncol(hembio)])
sw.pvalue <- as.numeric(unlist(sw.test))
sw.results <- data.frame(cols, sw.pvalue)
sw.results$Normality <- NA
sw.results <- as.data.table(sw.results)

sw.results$Normality <- sw.results[,apply()] # FINISH THIS. For values >0.05, write "NORMAL". Subset by the !NORMAL. Apply log(e). Test again. Write NORMAL or NON-NORMAL. Do the same for Levene "HOMOGENEOUS" - "NON-HOMOG". The variables that are NORMAL & HOMOGENEOUS, use parametric, the others non parametric, group by variables in col 2:4.
```

