---
title: "Tapirus terrestris: Health analysis - INFECTIOUS DISEASES"
author: "Tati Micheletti"
date: "16 May 2018"
output: html_document
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "Results") })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(data.table)
library(reproducible)
library(stats)
```


# Project Description

The present project aimed at understanding the general health of Tapir (**Tapirus terrestris**) populations on three different biomes.

# Diseases

## Analysis summary
diseases: only prevalence (X2)

1. Prevalences --> % positives on the population (relative prevalence). Use the positives in the total tested. (For each disease, positives/total ind. tested) Ex. 
A - disease1, disease2
B - Negative
C - disease1
D - disease1
E - Negative

Prevalence disease1: 3/5 = 0.6 or 60%
Prevalence disease2: 1/5 = 0.2 or 20%

To calculate the standard error of the binomial distribution, we have:

SE = sqrt(pH*(1-pH)/(n+1)), where pH = (x+1/2)/(n+1), `x` is the number of "positives" and `n` is the total number of samples 

## Coding

### 1. Load Data
Loading the data to work with from a `.csv` file and subsetting to diseases that have both positive and negative results.
```{r load}
infectiousFullAll <- data.table::fread(file.path(getwd(), "dataTapirus/Diseases/diseases.csv")) # loading from the dataTapirus
infectiousFull <-  infectiousFullAll[]#[biome == "CERRADO"]
  for (i in names(infectiousFull)) infectiousFull[get(i)=="", (i) := NA]
cols <- colnames(infectiousFull)
colsToKeep <- unlist(lapply(X = cols, FUN = function(columns){
  if ("Negativo" %in% infectiousFull[[columns]] &
      "REAGENTE" %in% infectiousFull[[columns]] &
      !"RESULTADO" %in% infectiousFull[[columns]])
    return(TRUE) else return(FALSE)
})
)
infectiousFull <- infectiousFull[, colsToKeep, with = FALSE] %>%
  cbind(infectiousFullAll[, "biome"], .)
infectiousFull

prevalTableTotalInfec <- infectiousFull[, lapply(.SD, function(x) sum(!is.na(x))), by = biome, .SDcols = names(infectiousFull)[names(infectiousFull) != "biome"]]

```

### 2. calculate frequencies of positives counts
```{r calculate_fre}
cols <- colnames(infectiousFull)[colnames(infectiousFull) != "biome"]
counts <- data.table::rbindlist(lapply(X = cols, FUN = function(dis){
    sb <- infectiousFull[, c("biome", ..dis)]
    tbl <- as.data.frame.matrix(table(sb))
    colnames(tbl)[which(names(tbl) == "Negativo")] <- "negative"
    colnames(tbl)[which(names(tbl) == "REAGENTE")] <- "freq"
    tbl$biome <- rownames(tbl)
    tbl$disease <- dis
    sb <- tbl
  if (!"ATLANTIC" %in% unique(sb$biome)){
    rowAT <- data.table::data.table(biome = "ATLANTIC",
                                    species = dis,
                                    freq = 0,
                                    negative = prevalTableTotalInfec[biome == "ATLANTIC", ..dis])
    sb <- rbind(sb, rowAT)
  }
    if (!"CERRADO" %in% unique(sb$biome)){
    rowCE <- data.table::data.table(biome = "CERRADO",
                                    species = dis,
                                    freq = 0,
                                    negative = prevalTableTotalInfec[biome == "CERRADO", ..dis])
    sb <- rbind(sb, rowCE)
    }
    if (!"PANTANAL" %in% unique(sb$biome)){
    rowPT <- data.table::data.table(biome = "PANTANAL",
                                    species = dis,
                                    freq = 0,
                                    negative = prevalTableTotalInfec[biome == "PANTANAL", ..dis])
    sb <- rbind(sb, rowPT)
    }
  return(sb)
})
)
``` 

### 3. Calculate the estimate (prevalence) and respective confidence intervals
``` {r ci}
CI <- data.table::rbindlist(lapply(X = cols, FUN = function(dis){
    biome.ci <- data.table::rbindlist(lapply(X = unique(counts$biome), FUN = function(biom){
  sb <- counts[biome == biom & disease == dis,]
  total <- sum(sb$freq, sb$negative)
  CI.calc <- stats::binom.test(x = sb$freq, n = total, p = 0.5,
                                   alternative="two.sided",
                                   conf.level=0.95)
  CI.low <- min(CI.calc$conf.int)
  CI.up <- max(CI.calc$conf.int)
  Ci <- data.table::data.table(biome = biom,
                               disease = dis,
                               prevalence = round(CI.calc$estimate, 3)*100,
                               CI = paste0(round(CI.low, 3)*100, " - ", round(CI.up, 3)*100))
  return(Ci)
})
)
    return(biome.ci)
})
)

CI

write.csv(x = CI, file = file.path(getwd(), "Results/infectiousPrevalence.csv"))
```

#### 4. X2 of the diseases' prevalence 
```{r prevalence}
biomeList <- infectiousFullAll[, "biome"]
  for (i in names(infectiousFullAll)) infectiousFullAll[get(i)=="", (i) := NA]
cols <- colnames(infectiousFullAll)
colsToKeep <- unlist(lapply(X = cols, FUN = function(columns){
  if ("Negativo" %in% infectiousFullAll[[columns]] &
      "REAGENTE" %in% infectiousFullAll[[columns]] & 
      !"RESULTADO" %in% infectiousFullAll[[columns]])
    return(TRUE) else return(FALSE)
})
)
infectiousFullAll <- infectiousFullAll[, colsToKeep, with = FALSE] %>%
  cbind(biomeList, .)

counts <- data.table::rbindlist(lapply(X = names(infectiousFullAll)[names(infectiousFullAll) != "biome"], 
                                       FUN = function(dis){
      countsByBiome <- data.table::rbindlist(lapply(X = unique(infectiousFullAll$biome), FUN = function(biom){
          sb <- infectiousFullAll[biome == biom, ]
          tble <- as.data.frame(table(sb[[dis]]), responseName = "freq")
          colnames(tble)[1] <- "result"
          tble$disease <- dis
          tble$biome <- biom
            if (all(length(unique(tble$result)) == 1 & 
              unique(tble$result) == "Negativo")) {
            rowREAGENTE <- data.table::data.table(result = "REAGENTE", 
                                                  freq = 0,
                                                  disease = dis,
                                                  biome = biom)
           tble <- rbind(tble, rowREAGENTE)
          }
           if (all(length(unique(tble$result)) == 1 &
               unique(tble$result) == "REAGENTE")) {
            rowNEGATIVO <- data.table::data.table(result = "Negativo", 
                                                  freq = 0,
                                                  disease = dis,
                                                  biome = biom)
            tble <- rbind(tble, rowNEGATIVO)
          }
          return(tble)
      })
      )
      return(countsByBiome)
})
)

countsCast <- data.table::dcast.data.table(data = counts, formula = biome + disease ~ result, value.var = "freq")
names(countsCast)[3:4] <- c("freq", "negative")

x2diseases <- data.table::rbindlist(lapply(X = unique(countsCast[, disease]), FUN = function(dis){
  sb <- countsCast[disease == dis, ]
  matrix <- as.matrix(sb[, c("freq", "negative")])
    rownames(matrix) <- sb$biome
  x2 <- rcompanion::pairwiseNominalIndependence(matrix, compare = "row", fisher = TRUE,
  gtest = FALSE, chisq = TRUE, method = "bonferroni", correct = "none",
  digits = 3)
  x2$species <- dis
  return(invisible(x2))
})
)

x2diseases
write.csv(x = x2diseases, file = file.path(getwd(), "Results/X2infectious.csv"))
```

