---
title: "Tapirus terrestris: Health analysis - MICROBIOLOGY"
author: "Tati Micheletti"
date: "May 15, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(reproducible)
library(rdrop2)
library(car)
library(dplyr)
library(plyr)
library(reshape2)
library(rcompanion)
library(betapart)
 token <- drop_auth()
```


# Project Description

The present project aimed at understanding the general health of Tapir (**Tapirus terrestris**) populations on three different biomes.

# Microbiology

## Analysis summary
bacteria: only prevalence (X2) and Jaccard

1. Prevalences --> % positives on the population. Use the positives in the total tested. (For each bacteria, positives/total ind. tested) Ex. 
A - bac1, bac2
B - Negative
C - bac1
D - bac1
E - Negative

Prevalence bac1: 3/5 = 0.6 or 60%
Prevalence bac2: 1/5 = 0.2 or 20%

2. Jaccard similarity coefficient for bacterias for cavities. Sj=1, same bio profile, Sj=0, different
     [Sj = j/(a+b-j)], j = number of species of bacteria that occurred in both the tapir populations for each cavity in the AF, PA (and CE?) study sites; a = number of species of bacteria that occurred in the PA population; b = number of species of bacteria that occurred in the AF population. AF and PA --> Sj = 0.125
3. x2 (chi-square) test for prevalence in different bacterias in study sites --> Not significant for the bacts found in both PT and MA

## Coding

### 1. Load Data

Loading the data to work with from a `.csv` file.
```{r load}
microbioFull <- data.table::fread("C:/Users/Tati/Dropbox/dataTapirus/microbioCerradoFull.csv")
microbioBinary <- data.table::fread("C:/Users/Tati/Dropbox/dataTapirus/microbioCerradoBinary.csv")
microbioPTAT <- data.table::fread("C:/Users/Tati/Dropbox/dataTapirus/PTATmicrobioFull.csv")
```
**SIDE NOTE: If the dataloading fails, turn on the VPN!.**

### 2. Prevalence calculation:

OBS.: Save the table to insert in the .xlsx make it a template for next captures

```{r columnsForPrevalence}
sumNonEmpty <- data.table::data.table(cavities = names(microbioBinary)[2:ncol(microbioBinary)], 
                                      numberSamples = apply(X = microbioBinary, MARGIN = 2, 
                                                            FUN = function(x)
                                                              sum(x != ""))[2:ncol(microbioBinary)])

tableCerrado <- data.table::rbindlist(lapply(X = names(microbioFull)[2:ncol(microbioFull)], FUN = function(x) {
  a <- plyr::count(microbioFull, x)
  a["cavity"] <- names(a)[1]
  a["biome"] <- "CERRADO"
  names(a) <- c("species", "frequency", "cavities", "biome")
  a <- a[!(a$species == ""),]
  return(a)
  })
)

tableCerrado$numberSamples <- apply(X = tableCerrado, MARGIN = 1, FUN = function(x) {
  num <- sumNonEmpty[cavities == x["cavities"], numberSamples]
  return(num)
  })


prevalenceMicrobio <- data.table::rbindlist(lapply(X = tableCerrado[, species], FUN = function(sp){
  sb <- tableCerrado[tableCerrado[, species == sp],]
  cavPrev <- data.table::rbindlist(lapply(X = sb[, cavities], FUN = function(cav){
    cavTable <- sb[sb[, cavities == cav]]
      total <- cavTable$numberSamples
      positive <- cavTable$frequency
  CI.calc <- stats::binom.test(x = positive, n = total, p = 0.5,
                                   alternative="two.sided",
                                   conf.level=0.95)
  CI.low <- min(CI.calc$conf.int)
  CI.up <- max(CI.calc$conf.int)

  Ci <- data.table::data.table(bacteria = sp,
                               cavity = cav,
                               prevalence = round(CI.calc$estimate, 3)*100,
                               CI.low = round(CI.low, 3)*100,
                               CI.up = round(CI.up, 3)*100,
                               CI.Interval = paste0(round(CI.low, 3)*100, " - ", round(CI.up, 3)*100))
    return(Ci)
  }))
  return(cavPrev)
})
)

tableCerrado$prevalence <- tableCerrado$frequency/tableCerrado$numberSamples*100
finalPrevalence <- rbind(microbioPTAT, tableCerrado)

prevalenceMicrobio

# write.csv(x = finalPrevalence, file = "C:/Users/Tati/Dropbox/dataTapirus/prevalenceMicrobio3AUG18.csv")
# Done already!
```

#### 3. X2 for the prevalence per cavity (we need to exclude cavities that were not accessed in PT and MA)
```{r X2_prevalence}
listX2 <- lapply(X = unique(finalPrevalence$species[finalPrevalence$species != "NEGATIVE"]), FUN = function(sp){
  assign(x = sp, value = finalPrevalence[species == sp, c("biome", "cavities", "frequency")])
  assign(x = sp, value = data.table::dcast(get(sp), biome ~ cavities, value.var = "frequency"))
  if (nrow(get(sp)) == 3){
    eval(parse(text = paste0("attributes(", sp, ")[['species']] <- sp")))
    return(get(sp))
  }
})

listX2 <- listX2[vapply(listX2, Negate(is.null), NA)]
listNames <- unlist(lapply(X = listX2, FUN = function(species){
 listName <- attributes(species)$species
  return(listName)
})
)
names(listX2) <- listNames

toExclude <- c("urethralSwab", "lesionSwab", "eyeSwab") # These were not collected (or compared) in the first paper, not much information
listX2 <- lapply(X = listX2, function(x){
  x[, !toExclude, with = FALSE]
  })

x2 <- lapply(X = listX2, FUN = function(x){
  for (i in names(x)) x[is.na(get(i)), (i) := 0]
  matrix <- as.matrix(x[,2:ncol(x)])
  rownames(matrix) <- x$biome
  #chisq.test(x = matrix, simulate.p.value = TRUE)
  pairwiseNominalIndependence(matrix, compare = "row", fisher = TRUE,
  gtest = FALSE, chisq = TRUE, method = "bonferroni", correct = "none",
  digits = 3)
})

X2pVal <- rbindlist(lapply(X = names(x2), FUN = function(species){
  x2[[species]]$species <- species
  return(x2[[species]])
})
)
X2pVal
```

In this table we can see that the p of both Fisher and X2 tests for E. coli are significant (Cerrado being different from Pantanal and MT Atlantica), but as we have to correct because of the low number of samples, they become non-significant. For `staphylococcusAureus` is not significant. 

#### 4. Similarity index (IF need to chose between PT and MT, do it with Pantanal)

Prepare the table for Jaccard: rows are sites, columns are species

```{r Jaccard2}
colsToKeep <- c("biome", "species")
dissimDF <- data.table::dcast(data = finalPrevalence[, colsToKeep, with = FALSE], formula = biome ~ species) %>%
  as.data.frame()
biomes <- dissimDF[, 1]
dissimDF <- dissimDF[, 3:ncol(dissimDF)]
rownames(dissimDF) <- biomes
dissimDF[dissimDF > 1] <- 1
AT.PT <- dissimDF[c("ATLANTIC", "PANTANAL"),]
CE.PT <- dissimDF[c("CERRADO", "PANTANAL"),]
CE.AT <- dissimDF[c("ATLANTIC", "CERRADO"),]

AT.PT.Sorensen <- betapart::beta.multi(x = AT.PT, index.family = "sorensen")
AT.PT.Jaccard <- betapart::beta.multi(x = AT.PT, index.family = "jaccard")

CE.PT.Sorensen <- betapart::beta.multi(x = CE.PT, index.family = "sorensen")
CE.PT.Jaccard <- betapart::beta.multi(x = CE.PT, index.family = "jaccard")

CE.AT.Sorensen <- betapart::beta.multi(x = CE.AT, index.family = "sorensen")
CE.AT.Jaccard <- betapart::beta.multi(x = CE.AT, index.family = "jaccard")

finalSimilarityTable <- data.table::data.table(
  Comparison = c("ATLANTIC:PANTANAL", "CERRADO:PANTANAL", "CERRADO:ATLANTIC"),
  Sorensen = c(AT.PT.Sorensen$beta.SOR, CE.PT.Sorensen$beta.SOR, CE.AT.Sorensen$beta.SOR),
  Jaccard = c(AT.PT.Jaccard$beta.JAC, CE.PT.Jaccard$beta.JAC, CE.AT.Jaccard$beta.JAC))

tdissimDF <- data.table::transpose(dissimDF)
colnames(tdissimDF) <- rownames(dissimDF)
rownames(tdissimDF) <- colnames(dissimDF)

tdissimDF
```

CONCLUSION: Similar MMATA ATLANTICA < CERRADO < PANTANAL (OH!!! >< )